---
title: "monocle tutorial"
author: "Emeric Texeraud"
date: "29/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Monocle
There are 2 means to use monocle. The first one is to use a Seurat object and use `SeuratWrappers` to turn it into a Cell dataset. The second one is to use only the count matrix and meta data, and use the monocle preprocessing method.

## Monocle with Seurat
Packages to install and use
```{r monocle_Seurat packages, eval=FALSE}
library(Seurat)
library(SeuratWrappers)
library(monocle3)
```

Seurat to cell dataset : 
```{r monocle_Seurat to_cell_dataset, eval=FALSE}
# load Seurat object
pal_alra <- readRDS('Pallial_ALRA_wDCs.rds')

cds <- as.cell_data_set(pal_alra)
```

learn graph :
```{r monocle_Seurat learn_graph, eval=FALSE}
cds <- cluster_cells(cds = cds, reduction_method = "UMAP")
cds <- learn_graph(cds, use_partition = FALSE)
plot_cells(cds,
           color_cells_by = "Seurat_clusters",
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE)
```

order cells : 
```{r monocle_Seurat order_cells, eval=FALSE}
cds <- order_cells(cds)

plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=TRUE,
           label_branch_points=FALSE,
           graph_label_size=1.5, cell_size = 1)
```

from cell dataset to Seurat
```{r monocle_Seurat to_seurat, eval=FALSE}
pal_monocle <- as.Seurat(cds)
```



## Monocle using count matrix
Packages to install and use
```{r monocle_Seurat packages, eval=FALSE}
library(Seurat) # optional, can be something else if you're using another package for your data
library(monocle3)
```

Create cell dataset :
```{r monocle_solo create_data, eval=FALSE}
# load data
pal_alra <- readRDS('Pallial_ALRA_wDCs.rds')

expression_matrix <- pal_alra@assays$RNA@counts
cell_metadata <- pal_alra@meta.data
gene_annotation <- data.frame(gene_short_name = row.names(expression_matrix), row.names = row.names(expression_matrix))

cds <- new_cell_data_set(expression_matrix,
                         cell_metadata = cell_metadata,
                         gene_metadata = gene_annotation)
```

preprocess the data with monocle : 
```{r monocle_solo preprocess, eval=FALSE}
cds <- preprocess_cds(cds, num_dim = 20, verbose = TRUE, norm_method = "log") # normalization only if raw counts

cds = align_cds(cds, num_dim = 20, alignment_group = "Condition") #To remove batch effects between Nx1 and Nx2, only if potential batch effect
```

reduce dimensionality : 
```{r monocle_solo reduce_dim, eval=FALSE}
# Reduce dimensionality and visualize the cells
cds <- reduce_dimension(cds, umap.min_dist = 0.3, umap.n_neighbors = 30L, verbose = TRUE, preprocess_method = "Aligned")
plot_cells(cds, color_cells_by = "Seurat_clusters", cell_size = 1)
```

learn graph :
```{r monocle_solo learn_graph, eval=FALSE}
cds <- cluster_cells(cds = cds, reduction_method = "UMAP")
cds <- learn_graph(cds, use_partition = FALSE)
plot_cells(cds,
           color_cells_by = "Seurat_clusters",
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE)
```

order cells : 
```{r monocle_Seurat order_cells, eval=FALSE}
cds <- order_cells(cds)

plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=TRUE,
           label_branch_points=FALSE,
           graph_label_size=1.5, cell_size = 1)
```

Then two options to extract the pseudotime. Either transform the cell dataset object in Seurat object, or just extract the pseudotime information and add it to your object.

Transform cell dataset to Seurat : 
```{r monocle_solo to_seurat, eval=FALSE}
pal_monocle <- as.Seurat(cds)
```

Extract pseudotime: 
```{r monocle_solo extract_pseudotime, eval=FALSE}
traj.plot <- plot_cells(pal_cds,
                        color_cells_by = "pseudotime",
                        label_cell_groups=FALSE,
                        label_leaves=TRUE,
                        label_branch_points=FALSE,
                        graph_label_size=1.5, cell_size = 1)

pal_alra@meta.data$monocle3_pseudotime <- traj.plot[["data"]][["cell_color"]]
```

## Visualize pseudotime with Seurat
```{r monocle visualize_pseudotime, eval=FALSE}
FeaturePlot(sub_sub_monocle, "monocle3_pseudotime", cols = c("darkblue", "yellow"))
```