---
title: "Exporting Seurat data for velocity"
author: "[LabEx CORTEX Bioinformatics Platform - Lyon 1 University -](https://labex-cortex.universite-lyon.fr/) <br>Vignette produced by Emeric Texeraud"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
---

The velocity analysis will be done in Python, and will not use the counts matrix from the Seurat object. Instead, the spliced and unspliced counts computed by velocyto will be used. But from these matrix, we will need to keep only the cells of interest, and keep the clusters and UMAP used throughout the analysis. Thus we will export these from our object, and use them for the velocity aanalysis.

## Load the Seurat object
```{r, warning=FALSE, include=TRUE}
# Load packages
suppressMessages({
  library(Seurat)
})

# Load the dataset
seur <- readRDS("/mnt/Data1/public_vignettes/seurat_data/p12_new.rds")
```

## Export data
```{r}
setwd('/mnt/Data1/public_vignettes/scvelo_data')
write.csv(x = seur@meta.data, file = 'seur_meta.csv')
write.csv(x = seur@reductions$umap@cell.embeddings, file = 'seur_umap.csv')
```







