---
title: "Scvelo using reticulate tutorial"
author: "[LabEx CORTEX Bioinformatics Platform](https://labex-cortex.universite-lyon.fr/) - Lyon 1 University - vignette produced by Emeric Texeraud"
date: "`r Sys.Date()`"
output: html_notebook
---

## Installing python and scvelo
For this first part, every command must be done in command line. The lines does not depend on your operating system, they will be the same for Linux, MacOs or Windows.

Installation of conda or python however may defer depending on the operating system, please refer to their respectives websites.

### Method 1 using base python

You can install python by clicking on this [link](https://www.python.org/downloads/)

To install scvelo on the base environnement, you can do as follow.
```{bash base intall scvelo}
# bash 
pip install -U scvelo
```

To verify scvelo installation, do as follow. If it throws an error, then scvelo was not properly installed
```{bash base verify installation}
# bash
python
>>> import scvelo as scv
```

### Method 2 using conda

You can install conda by clicking on this [link](https://conda.io/projects/conda/en/latest/user-guide/install/index.html)

You need to create a conda environment before installing scvelo. You may specify the python version you want, currently the most recent one is 3.9.
```{bash conda create env}
# bash
# create env named r-scvelo
conda create --name r-scvelo python=3.9 anaconda
```

Then you activate the newly created environment, and install scvelo. 
```{bash conda install scvelo}
# bash
# install and/or upgrade scvelo last version
conda activate r-scvelo
pip install -U scvelo
```

You can verify scvelo installation as follow. If it throws an error, then scvelo was not properly installed
```{bash conda verify installation}
# bash
# verify installation
conda activate r-scvelo
python
>>> import scvelo as scv
```

## Setting up reticulate and libraries
```{r libraries, warning=FALSE, message=FALSE}
library(reticulate)
library(Seurat, quietly = TRUE) # We need Seurat to access metadata of Seurat object
```

You need to know where conda is located in order for reticulate to find it.
```{bash}
# bash
which conda
```

then you just create a variable containing conda path, you will need to add the `conda = conda_path` argument for each reticulate command calling conda.
```{r}
conda_path = '/home/labex-bioinfo/python/anaconda3/condabin/conda'
```


You can get a list of conda libraries, if you forgot the environment name.
```{r conda list, warning=FALSE, message=FALSE}
conda_list(conda = conda_path)
```

You need to to tell reticulate which python environment is needed.
```{r reticulate set env}
use_condaenv("r-scvelo", required = TRUE, conda = conda_path)
```

Now we import all needed libraries from conda env. Even if you have an error about XMLRPC request failing, if you manage to get the message saying your scvelo and python version, then the import was successful.
```{python}
import scvelo as scv
import pandas as pd
import anndata
scv.logging.print_version()
```

## Loading all data
```{python}
# python chunk, generation of adata file is not complete when using in R chunk
# Load loom file generated with velocyto
adata = scv.read_loom(filename='/mnt/Data1/public_vignettes/cellranger_output/neuron_1k/velocyto/neuron_1k.loom')
adata.var_names_make_unique()
temp_X = adata._X
```
```{r}
# Load seurat analysed dataset and extract UMAP and metadata informations
seur <- readRDS("/mnt/Data1/public_vignettes/seurat_data/neuron_1k.rds")
meta_data <- seur@meta.data
umap_coordinates <- as.data.frame(seur@reductions$umap@cell.embeddings, row.names = Cells(seur))
```

```{r}
# Check cell names of adata and seurat object
adata$obs_names[1]
Cells(seur)[1]
```
If cell names have the same prefix and suffix, then you don't need to rename them. But if these are different, you need to rename the cells.

The function rename_cells takes the cell names of adata object as parameter. Then the prefix and suffix to remove from them, and then the prefix and suffix to add to the bare barcode to get the same cell names as the Seurat cell names.
```{r}
rename_cells <- function(cells, trim_start='', trim_end='', add_start='', add_end='') {
  paste0(add_start, gsub(paste0("^", trim_start, "|", trim_end, "$"), "", cells), add_end)}
renamed_cells = rename_cells(py$adata$obs_names$to_list(), trim_start = "neuron_1k:", trim_end = "x", add_end = "-1")
```

```{r}
umap_coordinates_py = r_to_py(umap_coordinates)
meta_data_py = r_to_py(meta_data)
renamed_cells_py = r_to_py(renamed_cells)
```


```{python}
# python chunk, because of errors when trying to subset it in r chunk
adata.obs_names = r.renamed_cells_py
cells = list(r.umap_coordinates_py.index)
adata._X = temp_X
adata = adata[cells]
adata.obsm['X_umap'] = r.umap_coordinates_py[adata.obs_names.to_list()].values
adata.obs['clusters'] = r.meta_data_py['clusters'].loc[adata.obs_names.to_list()].values)
```

```{r}
adata_py = ad$AnnData(obs = r_to_py(adata$obs), var = r_to_py(adata$var), obsm = r_to_py(adata$X), layers = r_to_py(adata$layers))
```


```{r}

scv$pl$proportions(adata_py, groupby = 'clusters')
```

