---
title: "Analysing cell cycle"
author: "[LabEx CORTEX Bioinformatics Platform - Lyon 1 University -](https://labex-cortex.universite-lyon.fr/) <br>Vignette produced by Emeric Texeraud"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
---

```{r load packages, warning=FALSE, include=TRUE}
# Load packages
suppressMessages({
  library(tidyverse)
  library(Seurat)
  library(gridExtra)
})

seur <- readRDS("/mnt/Data1/public_vignettes/seurat_data/p12_new.rds")
```

# Cell cycle analysis

To find our cycling cells, we first need to get the genes involved in the G2M and S phases of the cell cycle. For that, we will get the human cycling genes from seurat, and transform them to mouse format (uppercase first letter).
```{r cell cycle genes}
s.genes <- str_to_title(cc.genes.updated.2019$s.genes)
g2m.genes <- str_to_title(cc.genes.updated.2019$g2m.genes)
```

We then use the `CellcycleScoring` function from Seurat to score the S and G2M phases, and assign them to cells. What it does is scoring the S phase using S genes, then scoring the G2M phase using G2M genes. Finally, it will assign the cells to S phase if they have a high score for S genes, or G2M for a high G2M score.
```{r cell cycle scoring}
seur <- CellCycleScoring(seur, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE)
```

We can then see which cells have a high S score or high G2M score.
```{r, fig.width=8, fig.asp=0.4}
FeaturePlot(seur, features = c("S.Score", "G2M.Score"))
```
We can also see which cells are assigned S or G2M score.
```{r}
DimPlot(seur, group.by = "Phase")
```


